---
title: "CTAB DNA extraction"
author: "Ryan Eckert -- ryan.j.eckert@gmail.com"
output:
  html_document:
    theme: darkly
    toc: yes
    toc_depth: 3
    toc_float: yes
    css: "../ryan.css"
---
<link rel="stylesheet" href="../dracula.css">
<script src="http://cdnjs.cloudflare.com/ajax/libs/highlight.js/9.4.0/highlight.min.js"></script>

<a href="https://github.com/RyanEckert" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#dab844; color:#282a36; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>

```{r, setup, include = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE)

if (!require("pacman")) install.packages("pacman")

pacman::p_load(kableExtra)

supplies = read.csv("supplies.csv")
ctab = read.csv("ctab.csv")
reagents = read.csv("reagents.csv")
```
#### version: `r Sys.Date() %>% format(format="%B %d, %Y")`  

***
# About this document
This protocol is used to extract high quality genomic DNA from both coral host and Symbiodiniaceae spp. The resulting DNA should be sufficient for downstream PCR and high-thoughout sequencing. This protocol was utilized to examine Symbiodiniaceae community structure across depth through *ITS2* metabarcoding in [Eckert RJ, Reaume AM, Sturm AB, Studivan MS and Voss JD (2020) Depth influences Symbiodiniaceae associations among *Montastraea cavernosa* corals on the Belize Barrier Reef. *Front. Microbiol.* 11:518. doi: 10.3389/fmicb.2020.00518](https://doi.org/10.3389/fmicb.2020.00518){target="_blank"}. 
    
Additionally, the DNA extraction and clean-up were used in [Eckert, R. J., Studivan, M. S., and Voss, J. D. (2019). Populations of the coral species *Montastraea cavernosa* on the Belize Barrier Reef lack vertical connectivity. *Sci. Rep.* 9, 7200. doi:10.1038/s41598-019-43479-x ](https://doi.org/10.1038/s41598-019-43479-x){target="_blank"} to examine coral host population genetic structure.<br><br>

# CTAB gDNA Extraction Protocol 
***
Revised from [Mieog et al. (2009)](https://doi.org/10.1111/j.1755-0998.2008.02222.x){target="_blank"} and [http://ccb.ucr.edu/lab_protocols.html](http://ccb.ucr.edu/lab_protocols.html){target="_blank"}
<br>

## Supplies
The following supplies are necessary to perform a round of 24 extractions.  

```{r, supplies table}
kable(supplies) %>%
  kable_styling(bootstrap_options = c("condensed", "hover","striped"), full_width = FALSE, 
                position = 'left') %>%
  column_spec(1, width = "16em") %>%
  column_spec(2, width = "18em") %>%
  column_spec(3, width = "10em") %>%
  row_spec(2, hline_after = TRUE)
```

## Reagent recipes
Recipes for making ectraction buffer and associated stock solutions necessary for CTAB gDNA extraction follow.
<br>

### 2X CTAB extraction buffer
Use heat and stirring to dissolve CTAB into solution before adding NaCl. Make CTAB buffer just prior to extractions. Can keep at room temperature for several days.
```{r, ctab buffer, fig.align = 'left'}
kable(ctab, col.names = (c("Reagent","per 20 mL", "Target"))) %>%
  kable_styling(bootstrap_options = c("condensed", "hover","striped"), full_width = 
                FALSE, position = 'left') %>%
  column_spec(1, width = "33em") %>%
  column_spec(2, width = "10em")
```
  
### Reagent stock recipes             
```{r, reagent stocks}
kable(reagents, col.names = (c("Reagent","per 100 mL"))) %>%
    kable_styling(bootstrap_options = c("condensed", "hover","striped"), full_width 
                  = FALSE, position = 'left') %>%
  column_spec(1, width = "13em") %>%
  column_spec(2, width = "27em")
```
 
## Genomic DNA Extraction  
Set heatblock to 55 ºC  
Set refridgerated centrifuge and set to 4 ºC  
  
1. Prepare CTAB extraction buffer just prior to use, do not add proteinase K.
2. Scrape tissue from coral fragment and place into a 2 mL tube with 0.1 mL (~ 0.075 g) of 0.5 mm glass beads.
3. Add 800 µL CTAB extraction buffer.
4. Add 0.8 µL Proteinase K. Seal tubes with parafilm. Invert to mix. 
5. Bead beat for 2--3 mins (6 M/s, three 45 sec intervals w/ 2 min cool down between).
6. Incubate at 60 ºC for 90 min while mixing.
7. Add 800 µL Chloroform:Isoamyl  Alcohol (24:1). Invert to mix. 
8. Mix and centrifuge at 20,000 x *g* for 15 mins at 4 ºC.
9. Transfer aqueous phase to new 2 mL tube (600 µL then 150 µL), taking care not to disturb interphase layer. 
10. Add 800 µL cold (-20 ºC) Isopropanol.
11. Mix and incubate for 20 min at -20 ºC.
12. Centrifuge at 20,000 x *g* for 20 min at 4 ºC.
13. Carefully pour off supernatant. 
14. Add 150 µL of 70% Ethanol at room temperature. Invert to mix.
15. Centrifuge at 20,000 x *g* for 5 min at 4 ºC.
16. Remove supernatant (pour off, quick spin, pipette off remaining avoiding pellet) and dry inverted on Kimwipe for 15 min at room temperature. 
17. Elute in 100--200 µL of 1X TE pH 8.0.
18. Incubate at 55 ºC for 10 min.
<br><br>

# Cleaning genomic DNA
***
After extracting genomic DNA, use Zymo DNA Clean & Concentrator-5 (D4014) to clean DNA and remove inhibitors prior to running a PCR. This greatly improves nanodrop 230/260 and 260/280 readings, dramatically increasing amplification success.  

**Prior to first use add ethanol to Wash Buffer and label bottle.**  

1. Measure DNA concentration (Nanodrop okay for this step). Prepare 0.6 mL tubes with 5µg (or less) DNA in 100 µL total volume 1X TE to be cleaned.
2. Set Elution Buffer for elution step in heat block at 60--70 ºC.
3. In a 2 mL tube add 2:1 volume of Binding Buffer:DNA (200 µL) to each volume of genomic 
DNA and vortex thoroughly.
4. Transfer the mixture to a provided Zymo-Spin Column in a collection tube.
5. Centrifuge 10,000 x *g* for 30 sec at room temperature. Discard flow through.
6. Add 200μL DNA Wash Buffer to the column. Centrifuge at 10,000 x *g* for 1 min at room 
temperature. Repeat.
7. Transfer the column to a new labeled 1.5 mL tube. Elute by adding 20 µL of Elution 
Buffer directly to the column matrix and incubate at room temperature for 3–5 min. 
Centrifuge for 30 seconds to elute DNA.
8. Check DNA quality with Nanodrop and quantify DNA flourescently (e.g. Qubit) and prepare 10 ng/µL dilutions.
<br><br>